---
title: "Returns Distributions"
author: "Ben Gilbert"
date: "9/1/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Examples of Characterizing Returns Distributions

* Skills covered:

  - set working directory
  - compute summary statistics with fBasics package
  - t-tests: test mean return, skewness and kurtosis
  - Jarque-Bera (JB) test
  - Use cbind() function to create new data frame, give column names

***

### Define directory and load packages

Load (and install if necessary) any packages that we want to use. If we want to define a working directory for this session, we can. 

```{r pckges, eval=FALSE, message=FALSE}
# install.packages("fBasics") # Remove the # sign to install this package if you 
                              # have not already. Does a lot of basic stats.
library(fBasics)              # tell R to use this package in this session.
# install.packages("quantmod")
library(quantmod)

# Alternative way to load packages: use the pacman package:
# if(!require(pacman))install.packages("pacman")
# pacman::p_load(fBasics,quantmod)

# set working directory, notice slashes go forward
setwd("C:/Users/bgilbert_a/Dropbox/Econometrics/TimeSeriesCourse") 
```

```{r echo=FALSE, message=FALSE}
# install.packages("fBasics") # Remove the # sign to install this package if you 
                              # have not already. Does a lot of basic stats.
library(fBasics)              # tell R to use this package in this session.
# install.packages("quantmod")
library(quantmod)

# Alternative way to load packages: use the pacman package:
# if(!require(pacman))install.packages("pacman")
# pacman::p_load(fBasics,quantmod)

# set working directory, notice slashes go forward
setwd("C:/Users/bgilbert_a/Dropbox/Econometrics/TimeSeriesCourse") 
```

***

### Pull Data from Yahoo!Finance and Investigate Returns

* First investigate some market indices

```{r}
getSymbols("XLE",from="2000-01-03")
# XLE is symbol for the SPDR Energy ETF

dim(XLE) # see dimensions of dataset
head(XLE) #see first six rows
chartSeries(XLE,type=c("line"),theme="white",TA=NULL)

# calculate various types of returns
XLE.dsrtn = 100*dailyReturn(XLE,leading=FALSE,type='arithmetic') # daily simple return
XLE.dlrtn = 100*dailyReturn(XLE,leading=FALSE,type='log') # daily log return
XLE.msrtn = 100*monthlyReturn(XLE,leading=FALSE,type='arithmetic') # monthly simple return
XLE.mlrtn = 100*monthlyReturn(XLE,leading=FALSE,type='log') # monthly log return

chartSeries(XLE.dlrtn,type=c("line"),theme="white",TA=NULL)
chartSeries(XLE.mlrtn,type=c("line"),theme="white",TA=NULL)

getSymbols("SPY",from="2000-01-03") 
# SPY is the symbol for the S&P 500 Index
dim(SPY) # see dimensions of dataset
head(SPY) #see first six rows
SPY.dsrtn = 100*dailyReturn(SPY,leading=FALSE,type='arithmetic')
SPY.dlrtn = 100*dailyReturn(SPY,leading=FALSE,type='log')
SPY.msrtn = 100*monthlyReturn(SPY,leading=FALSE,type='arithmetic')
SPY.mlrtn = 100*monthlyReturn(SPY,leading=FALSE,type='log')
```

* Now investigate some individual equities

  - SRE (Sempra), BRK (Berkshire Hathaway), DUK (Duke Energy)

```{r}
getSymbols("SRE",from="2000-01-03") 
dim(SRE) # see dimensions of dataset
head(SRE) #see first six rows

chartSeries(SRE,type=c("line"),theme="white",TA=NULL)
SRE.dsrtn = 100*dailyReturn(SRE,leading=FALSE,type='arithmetic')
SRE.dlrtn = 100*dailyReturn(SRE,leading=FALSE,type='log')
SRE.msrtn = 100*monthlyReturn(SRE,leading=FALSE,type='arithmetic')
SRE.mlrtn = 100*monthlyReturn(SRE,leading=FALSE,type='log')

chartSeries(SRE.dlrtn,type=c("line"),theme="white",TA=NULL)
chartSeries(SRE.mlrtn,type=c("line"),theme="white",TA=NULL)

# Note the quotes around "BRK-B" and `BRK-B`. Dash in name requires this. 
getSymbols("BRK-B",from="2000-01-03")  
dim(`BRK-B`) 
# notice that we have to use bracket quotes ` ` because of the dash in BRK-B
head(`BRK-B`) 
BRK.dsrtn = 100*dailyReturn(`BRK-B`,leading=FALSE,type='arithmetic')
BRK.dlrtn = 100*dailyReturn(`BRK-B`,leading=FALSE,type='log')
BRK.msrtn = 100*monthlyReturn(`BRK-B`,leading=FALSE,type='arithmetic')
BRK.mlrtn = 100*monthlyReturn(`BRK-B`,leading=FALSE,type='log')

getSymbols("DUK",from="2000-01-03") 
dim(DUK) # see dimensions of dataset
head(DUK) #see first six rows
DUK.dsrtn = 100*dailyReturn(DUK,leading=FALSE,type='arithmetic')
DUK.dlrtn = 100*dailyReturn(DUK,leading=FALSE,type='log')
DUK.msrtn = 100*monthlyReturn(DUK,leading=FALSE,type='arithmetic')
DUK.mlrtn = 100*monthlyReturn(DUK,leading=FALSE,type='log')
```

#### Create a data frame with each return series

* Daily simple, Daily log, Monthly simple, Monthly log

```{r}
# use cbind() for column bind. 
# Not exactly a merge - assumes rows correspond

# Daily simple returns
dsrtn = cbind(SPY.dsrtn,XLE.dsrtn,SRE.dsrtn,BRK.dsrtn,DUK.dsrtn) 
colnames(dsrtn) = c("SPY","XLE","SRE","BRK","DUK") # add variable names

# Daily log returns
dlrtn = cbind(SPY.dlrtn,XLE.dlrtn,SRE.dlrtn,BRK.dlrtn,DUK.dlrtn)
colnames(dlrtn) = c("SPY","XLE","SRE","BRK","DUK")

# Monthly simple returns
msrtn = cbind(SPY.msrtn,XLE.msrtn,SRE.msrtn,BRK.msrtn,DUK.msrtn)
colnames(msrtn) = c("SPY","XLE","SRE","BRK","DUK")

# Monthly log returns
mlrtn = cbind(SPY.mlrtn,XLE.mlrtn,SRE.mlrtn,BRK.mlrtn,DUK.mlrtn)
colnames(mlrtn) = c("SPY","XLE","SRE","BRK","DUK")
```

#### Calculate summary statistics for each return series

```{r}
basicStats(dsrtn) # calculate summary statistics from the fBasics package
basicStats(dlrtn)
basicStats(msrtn)
basicStats(mlrtn)
```

* Comparison to observations in Tsay Chapter 1:

  - daily have higher excess kurtosis than monthly. In monthly, market indices have higher excess kurtosis than individual stocks 
  
    - holds for energy
    
  - means of dailys are near zero while means of monthlies are higher
  
    - holds for energy
    
  - standard deviations of monthlies are also higher
  
    - holds for energy
    
  - In dailies, market indices have smaller standard deviations than individual stocks
  
    - not necessarily true for energy
  
  - skewness is not a big issue for any of them
  
    - holds for energy
    
  - simple and log returns are not that much different
  
    - holds for energy

***

### Testing hypotheses about returns distributions

* Example with daily Berkshire Hathaway log returns

#### Summary statistics

```{r}
# transform from zoo object to data frame
brkrt= na.omit(as.data.frame(BRK.dlrtn))

# Two different ways to calculate summary statistics
basicStats(brkrt)
mean(brkrt[,1])
var(brkrt[,1])
sqrt(var(brkrt[,1]))
skewness(brkrt[,1])
kurtosis(brkrt[,1])
```

#### t-test of mean returns

* several possible hypotheses

$$
H_0: \mu = 0, \;\; H_a: \mu \neq 0
$$

```{r}
t.test(brkrt[,1]) # t test of mean = 0
```

$$
H_0: \mu = 0, \;\; H_a: \mu > 0
$$
```{r}
t.test(brkrt[,1],alternative=c("greater")) # one-sided test
```

$$
H_0: \mu = 0.04, \;\; H_a: \mu \neq 0.04
$$
```{r}
t.test(brkrt[,1],mu=0.04) # two-sided test
```

$$
H_0: \mu = 0.04, \;\; H_a: \mu < 0.04
$$

```{r}
t.test(brkrt[,1],mu=0.04,alternative=c("less")) # one-sided test
```

#### Compare returns to normal density

* Plot histogram and density of returns vs. Normal distribution

```{r}
hist(brkrt[,1],nclass=40) # histogram of simple returns to compare to normal
d1 = density(brkrt[,1])   # fitted density, creates x-axis and y-axis
names(d1)

# plot the density
plot(d1$x,d1$y,type="l") # type is the letter l for line, not the numeral 1

# construct values for Normal density
mu = mean(brkrt[,1]); s1 = sd(brkrt[,1])
x = seq(-5,5,0.01)
y = dnorm(x,mean=mu,sd=s1)
lines(x,y,lty=2) # imposes a normal density over the empirical one for comparison
```

#### Statistical tests for Normality of returns

* Individual t-tests for skewness and kurtosis
* Joint Jarque-Bera (JB) test for both simultaneously

```{r}
# JB test
normalTest(brkrt[,1],method='jb') # Jarque-Bera test of normality

# Individual t-tests: skewness
s1=skewness(brkrt[,1])
T <- length(brkrt[,1]) # calculate time dimension/sample size
tst <- s1/sqrt(6/T) # test of skewness
tst
pv <- 2*(pnorm(tst)) # p value of test
pv

# Individual t-tests: kurtosis
k4 <- kurtosis(brkrt[,1])
tst <- k4/sqrt(24/T) # test of excess kurtosis
tst
```

***

### Replicating Examples from Tsay's Book

* Use read.table() function to read in a text file organized in data columns

```{r}
setwd("C:/Users/bgilbert_a/Dropbox/Econometrics/TimeSeriesCourse") 

# header=TRUE denotes variable names in first row
da=read.table('Fall2017/TsayCh1/Lecture/d-ibm3dx7008.txt',header=T) 

# daily simple returns of market Value Weighted, Equal Weighted, 
# and S&P Portfolio indices plus IBM.
# this data can replicate part of Tsay Table 1.2
head(da)
tail(da)
dim(da)
ibm = da[,2]
# time plot, could change title (main)
ts.plot(ibm,main="Daily simple returns of IBM") 
```

* Calculate summary statistics and t-tests

```{r}
sibm= ibm
basicStats(sibm)
mean(sibm)
var(sibm)
sqrt(var(sibm))
skewness(sibm)
kurtosis(sibm)

# t test of zero mean returns
t.test(sibm) 
t.test(sibm,alternative=c("greater")) # one-sided test
```

* Compare returns to normal density

```{r}
hist(sibm,nclass=40) # histogram of simple returns to compare to normal
d1 = density(sibm)
names(d1)
plot(d1$x,d1$y,type="l") # type is the letter l for line, not the numeral 1
mu = mean(sibm); s1 = sd(sibm)
x = seq(-0.4,0.4,0.01)
y = dnorm(x,mean=mu,sd=s1)
lines(x,y,lty=2) # imposes a normal density over the empirical one for comparison
```

* Test for Normality

```{r}
normalTest(sibm,method='jb') # Jarque-Bera test of normality
s1=skewness(sibm)
T <- length(sibm)   # calculate time dimension/sample size
tst <- s1/sqrt(6/T) # test of skewness
tst
pv <- 2*(pnorm(tst)) # p value of test
pv
k4 <- kurtosis(sibm)
tst <- k4/sqrt(24/T) # test of excess kurtosis
tst
```
    
* Same exercise for log returns instead of simple returns

```{r}
lnibm <- 100*log(ibm+1) # computes log return in percentage
basicStats(lnibm)
mean(lnibm)
var(lnibm)
skewness(lnibm)
kurtosis(lnibm)
basicStats(lnibm)
t.test(lnibm) # t test of mean = 0
t.test(lnibm,alternative=c("greater"))
# compare returns to normal density
hist(lnibm,nclass=40) # histogram of simple returns to compare to normal
d1 = density(lnibm)
names(d1)
plot(d1$x,d1$y,type="l") # type is the letter l for line, not the numeral 1
mu = mean(lnibm); s1 = sd(lnibm)
x = seq(-10,10,0.01)
y = dnorm(x,mean=mu,sd=s1)
lines(x,y,lty=2) # imposes a normal density over the empirical one for comparison

normalTest(lnibm,method='jb') # Jarque-Bera test of normality
s2=skewness(lnibm)
T <- length(lnibm) # calculate time dimension/sample size
tst <- s2/sqrt(6/T) # test of skewness
tst
pv <- 2*(pnorm(tst)) # p value of test
pv
k2 <- kurtosis(lnibm)
tst <- k2/sqrt(24/T) # test of excess kurtosis
tst
```

#### Visual examples

* Example from book with Apple stock

```{r}
setwd("C:/Users/bgilbert_a/Dropbox/Econometrics/TimeSeriesCourse") 

x=read.table("Tsay2017/Lecture1/d-aapl0413.txt",header=T) # Load Apple stock returns
dim(x) # check the size of the data file
x[1,] # show the first row of the data
y=ts(x[,3],frequency=252,start=c(2004,1)) # Create a time-series object in R.
plot(y,type='l',xlab='year',ylab='rtn') # type = letter l, not number 1
title(main='Daily returns of Apple stock: 2004 to 2013')
par(mfcol=c(2,1)) # To put two plots on a single page
y=y*100 # percentage returns
hist(y,nclass=50)
title(main='Percentage returns')
d1=density(y)
plot(d1$x,d1$y,xlab='returns',ylab='den',type='l')
```

* Example from book with T-bill spreads

```{r}
setwd("C:/Users/bgilbert_a/Dropbox/Econometrics/TimeSeriesCourse") 

x=read.table("Tsay2017/Lecture1/m-tb3ms.txt",header=T) # Load 3m-TB rates
dim(x)
y=read.table("Tsay2017/Lecture1/m-tb6ms.txt",header=T) # Load 6m-TB rates
dim(y)
914-615 # make sure dimensions of 3m and 6m match up
x[300,] # Check date of the 3m-TB
y[1,] # Check date of the 1st observation of 6m-TB
int=cbind(x[300:914,4],y[,4]) # Line up the two TB rates
tdx=(c(1:615)+11)/12+1959
par(mfcol=c(1,1))
max(int)
plot(tdx,int[,1],xlab='year',ylab='rate',type='l',ylim=c(0,16.5))
lines(tdx,int[,2],lty=2) # Plot the 6m-TB rate on the same frame.
plot(tdx,int[,2]-int[,1],xlab='year',ylab='spread',type='l')
abline(h=c(0)) # Draw a horizontal line to ``zero''
```

* Example from book with Coca-Cola corporate earnings

```{r}
setwd("C:/Users/bgilbert_a/Dropbox/Econometrics/TimeSeriesCourse") 

x=read.table("Tsay2017/Lecture1/q-ko-earns8309.txt",header=T) # Load KO data
dim(x)
x[1,]
tdx=c(1:107)/12+1983
plot(tdx,x[,3],xlab='year',ylab='earnings',type='l')
title(main='EPS of Coca Cola: 1983-2009')
points(tdx,x[,3])
```

* Example from book with US-EU exchange rates

```{r}
setwd("C:/Users/bgilbert_a/Dropbox/Econometrics/TimeSeriesCourse") 

y=read.table("Tsay2017/Lecture1/d-exuseu.txt",header=T) # Load USEU exchange rates
dim(y)
y[1,]
tdx=c(1:3567)/252+1999
plot(tdx,y[,4],xlab='year',ylab='eu',type='l')
title(main='Dollars per Euro')
r=diff(log(y[,4])) #= Compute log returns
plot(tdx[2:3567],r,xlab='year',ylab='rtn',type='l')
title(main='ln-rtn: US-EU')
hist(r,nclass=50)
title(main='useu: ln-rtn')
```

* Example from book with IBM returns

```{r}
setwd("C:/Users/bgilbert_a/Dropbox/Econometrics/TimeSeriesCourse") 

da=read.table('Tsay2017/Lecture1/m-ibm-6815.txt',header=T) # naming this da lets us run program with 
# different assets by changing at the top
head(da)
dim(da)
ibm=da$RET # simple return
lnIBM <- log(ibm+1) # computes log return
ts.plot(ibm,main="Monthly simple returns") # time plot, could change title (main)
# compare returns to normal density
hist(ibm,nclass=40) # histogram of simple returns to compare to normal
d1 = density(lnIBM)
names(d1)
plot(d1$x,d1$y,type="l") # type is the letter l for line, not the numeral 1
mu = mean(lnIBM); s1 = sd(lnIBM)
x = seq(-0.4,0.4,0.01)
y = dnorm(x,mean=mu,sd=s1)
lines(x,y,lty=2) # imposes a normal density over the empirical one for comparison
```
