---
title: "Returns Distributions"
author: "Ben Gilbert"
date: "9/1/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Returns Distributions Analysis

## Overview

* Skills covered:

  - Setting working directory
  - Computing summary statistics with fBasics package
  - Running t-tests for mean return, skewness and kurtosis
  - Performing Jarque-Bera (JB) test for normality
  - Using cbind() function to create new data frames
  - Visualizing return distributions

***

## 1. Environment Setup

### Loading Required Packages

Load (and install if necessary) any packages that we want to use. If we want to define a working directory for this session, we can. 

```{r load_packages, eval=FALSE, message=FALSE}
# install.packages("fBasics") # Uncomment to install if needed
library(fBasics)              # For statistical analysis
# install.packages("quantmod") # Uncomment to install if needed
library(quantmod)             # For financial data and returns calculation

# Alternative package loading approach using pacman
# if(!require(pacman))install.packages("pacman")
# pacman::p_load(fBasics, quantmod)

# Set working directory
setwd("C:/Users/gilbe/Dropbox/Econometrics/TimeSeriesCourse") 
```

```{r hidden_setup, echo=FALSE, message=FALSE}
# install.packages("fBasics") # Uncomment to install if needed
library(fBasics)              # For statistical analysis
# install.packages("quantmod") # Uncomment to install if needed
library(quantmod)             # For financial data and returns calculation

# Set working directory (adjust path as needed)
# setwd("C:/Users/gilbe/Dropbox/Econometrics/TimeSeriesCourse") 
```

***

## 2. Financial Data Analysis

### 2.1 Market Indices

First, we'll fetch and analyze market indices data using the quantmod package.

```{r market_indices}
# Create a function to plot price and return charts with proper formatting
plot_security_data = function(ticker_data, ticker_name) {
  chartSeries(ticker_data, type=c("line"), theme="white", TA=NULL,
             main=paste(ticker_name, "Price"))
}

# Fetch ETF representing energy sector
getSymbols("XLE", from="2000-01-03")
# XLE is symbol for the SPDR Energy ETF

# Examine data structure
dim(XLE)  # See dimensions of dataset
head(XLE) # See first six rows

# Visualize price data
plot_security_data(XLE, "SPDR Energy ETF (XLE)")

# Create a function to calculate different types of returns
calculate_returns = function(symbol_data) {
  returns = list(
    daily_simple = 100*dailyReturn(symbol_data, leading=FALSE, type='arithmetic'),
    daily_log = 100*dailyReturn(symbol_data, leading=FALSE, type='log'),
    monthly_simple = 100*monthlyReturn(symbol_data, leading=FALSE, type='arithmetic'),
    monthly_log = 100*monthlyReturn(symbol_data, leading=FALSE, type='log')
  )
  return(returns)
}

# Calculate returns for XLE
xle_returns = calculate_returns(XLE)

# Visualize returns
chartSeries(xle_returns$daily_log, type=c("line"), theme="white", TA=NULL,
           main="XLE Daily Log Returns")
chartSeries(xle_returns$monthly_log, type=c("line"), theme="white", TA=NULL,
           main="XLE Monthly Log Returns")

# Fetch S&P 500 Index ETF
getSymbols("SPY", from="2000-01-03") 
# SPY is the symbol for the S&P 500 Index
spy_returns = calculate_returns(SPY)
```

### 2.2 Individual Energy Stocks

Now we'll analyze individual energy company stocks:

- SRE (Sempra Energy)
- BRK-B (Berkshire Hathaway)
- DUK (Duke Energy)

```{r individual_stocks}
# Sempra Energy
getSymbols("SRE", from="2000-01-03") 
plot_security_data(SRE, "Sempra Energy (SRE)")
sre_returns = calculate_returns(SRE)

chartSeries(sre_returns$daily_log, type=c("line"), theme="white", TA=NULL,
           main="SRE Daily Log Returns")

# Berkshire Hathaway
# Note the quotes around "BRK-B" and `BRK-B`. Dash in name requires this. 
getSymbols("BRK-B", from="2000-01-03")  
# Must use backtick quotes because of the dash in BRK-B
brk_returns = calculate_returns(`BRK-B`)

# Duke Energy
getSymbols("DUK", from="2000-01-03") 
duk_returns = calculate_returns(DUK)
```

## 3. Combined Returns Analysis

### 3.1 Creating Return Dataframes

Let's combine our returns data into organized dataframes for comparison.

```{r combined_returns}
# Function to create combined returns dataframes
create_combined_returns = function() {
  # Daily simple returns
  daily_simple_returns = cbind(
    spy_returns$daily_simple,
    xle_returns$daily_simple,
    sre_returns$daily_simple,
    brk_returns$daily_simple,
    duk_returns$daily_simple
  )
  colnames(daily_simple_returns) = c("SPY", "XLE", "SRE", "BRK", "DUK")
  
  # Daily log returns
  daily_log_returns = cbind(
    spy_returns$daily_log,
    xle_returns$daily_log,
    sre_returns$daily_log,
    brk_returns$daily_log,
    duk_returns$daily_log
  )
  colnames(daily_log_returns) = c("SPY", "XLE", "SRE", "BRK", "DUK")
  
  # Monthly simple returns
  monthly_simple_returns = cbind(
    spy_returns$monthly_simple,
    xle_returns$monthly_simple,
    sre_returns$monthly_simple,
    brk_returns$monthly_simple,
    duk_returns$monthly_simple
  )
  colnames(monthly_simple_returns) = c("SPY", "XLE", "SRE", "BRK", "DUK")
  
  # Monthly log returns
  monthly_log_returns = cbind(
    spy_returns$monthly_log,
    xle_returns$monthly_log,
    sre_returns$monthly_log,
    brk_returns$monthly_log,
    duk_returns$monthly_log
  )
  colnames(monthly_log_returns) = c("SPY", "XLE", "SRE", "BRK", "DUK")
  
  return(list(
    daily_simple = daily_simple_returns,
    daily_log = daily_log_returns,
    monthly_simple = monthly_simple_returns,
    monthly_log = monthly_log_returns
  ))
}

# Create combined returns
combined_returns = create_combined_returns()
```

### 3.2 Calculating Summary Statistics

Let's calculate and analyze the summary statistics for each return series.

```{r summary_stats}
# Calculate summary statistics
basicStats(combined_returns$daily_simple)
basicStats(combined_returns$daily_log)
basicStats(combined_returns$monthly_simple)
basicStats(combined_returns$monthly_log)
```

### 3.3 Observations from the Data

Key observations from the returns data compared to Tsay Chapter 1:

- Daily returns have higher excess kurtosis than monthly returns
  - Monthly market indices have higher excess kurtosis than individual stocks
  - These patterns hold for energy stocks
  
- Means of daily returns are near zero, while monthly returns have higher means
  - This pattern holds for energy stocks
  
- Standard deviations of monthly returns are higher than daily
  - This pattern holds for energy stocks
  
- In Tsay, market indices have smaller standard deviations than individual stocks
  - This is not consistently true for energy stocks
  
- Skewness is not significant for most return series
  - This pattern holds for energy stocks
  
- Simple and log returns produce similar patterns
  - This pattern holds for energy stocks

## 4. Detailed Analysis of Berkshire Hathaway Returns

Let's perform a detailed analysis of Berkshire Hathaway's daily log returns.

```{r brk_analysis}
# Format BRK returns for analysis
brk_daily_log = na.omit(as.data.frame(brk_returns$daily_log))

# Summary statistics calculation function
calculate_stats = function(returns) {
  stats = list(
    mean = mean(returns),
    variance = var(returns),
    std_dev = sqrt(var(returns)),
    skewness = skewness(returns),
    kurtosis = kurtosis(returns)
  )
  return(stats)
}

# Calculate statistics
brk_stats = calculate_stats(brk_daily_log[,1])
print(brk_stats)

# Comprehensive stats from fBasics
basicStats(brk_daily_log)
```

### 4.1 Testing Mean Returns

We can test whether Berkshire Hathaway's mean returns differ significantly from zero.

```{r brk_t_tests}
# Two-sided test (H0: mean = 0, Ha: mean ≠ 0)
t.test(brk_daily_log[,1])

# One-sided test (H0: mean = 0, Ha: mean > 0)
t.test(brk_daily_log[,1], alternative=c("greater"))
```

### 4.2 Comparing Returns to Normal Distribution

Let's visualize how Berkshire's returns compare to a normal distribution.

```{r brk_distribution}
# Function to visualize return distribution vs normal
calculate_and_plot_density = function(returns, title="Returns Distribution") {
  # Create histogram
  hist(returns, nclass=40, main=title, xlab="Returns", ylab="Frequency")
  
  # Calculate and plot density
  density_estimate = density(returns)
  plot(density_estimate$x, density_estimate$y, type="l", 
       main=paste(title, "Density"), xlab="Returns", ylab="Density")
  
  # Add normal distribution overlay
  mean_val = mean(returns)
  sd_val = sd(returns)
  x_vals = seq(min(density_estimate$x), max(density_estimate$x), length.out=100)
  y_vals = dnorm(x_vals, mean=mean_val, sd=sd_val)
  lines(x_vals, y_vals, lty=2, col="red") 
  legend("topright", c("Empirical", "Normal"), lty=c(1,2), col=c("black", "red"))
}

# Plot distribution
calculate_and_plot_density(brk_daily_log[,1], "Berkshire Hathaway Daily Log Returns")
```

### 4.3 Testing for Normality

We can statistically test whether Berkshire's returns follow a normal distribution.

```{r brk_normality}
# Function to test normality
test_normality = function(returns, name="Returns") {
  cat("Normality Tests for", name, "\n")
  cat("------------------------------------------------\n")
  
  # Jarque-Bera test
  jb_test = normalTest(returns, method='jb')
  print(jb_test)
  
  # Test skewness
  s1 = skewness(returns)
  sample_size = length(returns)
  skew_test = s1/sqrt(6/sample_size)
  p_value_skew = 2*(pnorm(abs(skew_test), lower.tail=FALSE))
  cat("Skewness:", s1, "| Test statistic:", skew_test, "| p-value:", p_value_skew, "\n")
  
  # Test kurtosis
  k4 = kurtosis(returns)
  kurt_test = k4/sqrt(24/sample_size)
  cat("Kurtosis:", k4, "| Test statistic:", kurt_test, "\n")
  cat("------------------------------------------------\n\n")
}

# Run normality tests
test_normality(brk_daily_log[,1], "Berkshire Hathaway Daily Log Returns")
```

## 5. IBM Data Examples from Tsay

This section demonstrates analysis using IBM returns data provided in Tsay's textbook.

```{r tsay_ibm, eval=FALSE}
setwd("C:/Users/gilbe/Dropbox/Econometrics/TimeSeriesCourse") 
# Load IBM returns data
data_ibm = read.table('Fall2017/TsayCh1/Lecture/d-ibm3dx7008.txt', header=TRUE) 

# Extract IBM returns
ibm_returns = data_ibm[,2]

# Simple returns analysis
simple_returns = ibm_returns

# Log returns calculation
log_returns = 100*log(ibm_returns+1)

# Visualize distributions
calculate_and_plot_density(simple_returns, "IBM Simple Returns")
calculate_and_plot_density(log_returns, "IBM Log Returns") 

# Test normality
test_normality(simple_returns, "IBM Simple Returns")
test_normality(log_returns, "IBM Log Returns")
```

## 6. Additional Visualization Examples

```{r additional_examples, eval=FALSE}
setwd("C:/Users/gilbe/Dropbox/Econometrics/TimeSeriesCourse") 
# Apple stock returns example
x = read.table("Tsay2017/Lecture1/d-aapl0413.txt", header=TRUE)
y = ts(x[,3], frequency=252, start=c(2004,1))
plot(y, type='l', xlab='year', ylab='rtn',
     main='Daily returns of Apple stock: 2004 to 2013')

# T-bill spreads example
x = read.table("Tsay2017/Lecture1/m-tb3ms.txt", header=TRUE)
y = read.table("Tsay2017/Lecture1/m-tb6ms.txt", header=TRUE)
int = cbind(x[300:914,4], y[,4])
tdx = (c(1:615)+11)/12+1959

# Plot rates together with legend
plot(tdx, int[,1], xlab='year', ylab='rate', type='l', ylim=c(0,16.5),
     main="3-Month and 6-Month Treasury Bill Rates")
lines(tdx, int[,2], lty=2, col="blue")
legend("topright", c("3-Month", "6-Month"), lty=c(1,2), col=c("black", "blue"))

# Plot spread
plot(tdx, int[,2]-int[,1], xlab='year', ylab='spread', type='l',
     main="Spread between 6-Month and 3-Month T-Bills")
abline(h=c(0), col="gray")
```